---
title: "3+ Sample testing with continuous outcomes"

---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
```

## ANOVA


### Visualizing first

What graphical strategies have we learned to look at variability within and between groups?

### A) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
library(ggplot2)
library(tidyr)
library(dplyr)
set.seed(1)

group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 15, sd = 3)
group_3 <- rnorm(n = 10, mean = 15, sd = 3)

no_diff <- data.frame(group_1, group_2, group_3)

no_diff_narrow <- no_diff %>% gather(key = "Group", value = "Measure", group_1:group_3)

means <- no_diff_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(no_diff_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```


### Summary of the plots

Plot (A)

- The means (red dots) were not very different across the groups. This means
the variation **between** the group means was small.
- The distribution of the data (black Xs) was wide enough that the distribution 
of points for each group overlapped almost completely. This means that the variation
**within** each group was relatively wide 

### B) Is there a difference between these means?

Describe why you do or do not think so.


```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 20, sd = 3)
group_3 <- rnorm(n = 10, mean = 27, sd = 3)

diff <- data.frame(group_1, group_2, group_3)

diff_narrow <- diff %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_2 <- diff_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_2, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_2, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```


### Summary of the plots

Plot (B)

- The means are quite different across the groups. The variation **between** the
group means would be larger than in plot (A)
- The distribution of the data overlaps between groups 1 and 2 and 2 and 3, but 
not 1 and 3. The variation **within** each group is as wide as it was in Plot (A)
but doesn't mask the mean differences, especially between group 1 and 3


### C) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 20, sd = 3)
group_3 <- rnorm(n = 10, mean = 15, sd = 3)

diff_2 <- data.frame(group_1, group_2, group_3)

diff_2_narrow <- diff_2 %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_3 <- diff_2_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_2_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_3, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_3, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```


### Summary of the plots

Plot (C)

- Here, the means for group 1 and 3 look similar, but the mean for group 2 appears
a bit higher than the other two, though there is still overlap between the data 
from all the groups
- Is there evidence that at least one of the means is different?

### D) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 7)
group_2 <- rnorm(n = 10, mean = 20, sd = 7)
group_3 <- rnorm(n = 10, mean = 27, sd = 7)

diff_3 <- data.frame(group_1, group_2, group_3)

diff_3_narrow <- diff_3 %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_4 <- diff_3_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_3_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_4, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_4, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```


### Summary of the plots

Plot (D)

- Plot (D) looked like Plot (B) but with more variation **within** groups
- This variation makes the difference between the means harder to detect

### Overall summary

- What we informally did on the previous slides was compare the variation **between** 
group means to the variation **within** the groups
- This focus on variation is why this test is called ANOVA: an ANalysis Of VAriance
- When the ratio of between vs. within variation is large enough then we detect 
a difference between the groups
- When the ratio isn't large enough we don't detect the difference.
- This ratio is our test statistic, denoted by $F$

## Descriptive plots


### Box plot

```{r, out.width="70%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Group, y = Measure)) + 
  geom_boxplot() +
    geom_point() +
  theme_minimal(base_size = 15)
```


### Density plot

```{r, out.width="60%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Measure)) + 
  geom_density(aes(fill = Group), alpha = 0.5) +
  theme_minimal(base_size = 15)
```


### Histograms with facet wrap
```{r, out.width="60%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Measure)) + 
  geom_histogram(aes(fill = Group), col = "white", binwidth = 2.5) +
  theme_minimal(base_size = 15) + facet_wrap(~ Group, nrow = 3)
```

## Testing with ANOVA

### The hypotheses

**Null hypothesis**

$H_0: \mu_1=\mu_2=...\mu_K$, where $K$ is the number of levels of the grouping variable

- Can you also state the null hypothesis in words?


### The hypotheses
**Alternative hypothesis**

$H_a:$ not all $\mu_1$, $\mu_2$,..., $\mu_K$ are equal

- In words: Not all means are the same. Or, **at least one of the means** differs from the 
others.


### ANOVA in R: use `aov()`, then `tidy()` it up!

- `aov()` stands for analysis of variance

The general syntax for the ANOVA is:

aov(outcomevariable ~ groupvaraible, data=dataset)

We will save the output of this as an object and then use tidy(object) to get the output we want.

reference: https://broom.tidyverse.org/reference/anova_tidiers.html

### ANOVA in R: use `aov()`, then `tidy()` it up!
We will focus on two parts of the output from this package:

- `statistic` is the $F$ test statistic, the ratio of the variation between means
vs. the variation within groups.

- `p.value` is the p-value for the test. 

### p of an f statistic in R

You can check that you can calculate the p-value from the F distribution. Remember,
that you need to specify a degrees of freedom for the numerator and for the denominator:

pf(value, df1=numerator degrees of freedom, df2= denominator degrees of freedom, lower.tail=F)

This general pattern of syntax should look familiar by now....

