---
title: "Bonus material: permutations"

---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
library(readr)
library(ggplot2)
library(broom) 
library(tidyr)
library(ggplot2)
library(tibble)
library(infer)
```


### Example: Beer consumption and mosquito attraction to humans
Background: Malaria and alcohol consumption both represent major public health 
problems. Alcohol consumption is rising in developing countries and, as efforts 
to manage malaria are expanded, understanding the links between malaria and 
alcohol consumption becomes crucial. Our aim was to ascertain the effect of beer
consumption on human attractiveness to malaria mosquitoes in semi field 
conditions in Burkina Faso. - [Lefevre et al, 2010, in *PLOS One*](https://journals.plos.org/plosone/article?id=10.1371/journal.pone.0009546)

### Example: Beer consumption and mosquito attraction to humans

- Volunteers were randomly assigned to drink either beer or water
- Batches of mosquitoes were inside a device and could choose to fly towards the
human participant or towards the open air
- The number of mosquito flying towards the human were counted for each participant

### Example: Beer consumption and mosquito attraction to humans
The data:

```{r, echo=TRUE}
beer <- c(27, 19, 20, 20, 23, 17, 21, 24, 31, 26, 28, 20, 
          27, 19, 25, 31, 24, 28, 24, 29, 21, 21, 18, 27, 
          20)

water <- c(21, 19, 13, 22, 15, 22, 15, 22, 20, 12, 24, 24, 
           21, 19, 18, 16, 23, 20)

mosq_data <- data.frame(num_mosquitos = c(beer, water), 
                        treatment = c(rep("beer", 25),
                                      rep("water", 18)))
```


### Example: Beer consumption and mosquito attraction to humans
`beer` is a vector of the count of mosquitoes the flew towards the participant
for the 25 people randomized to beer

`water` is the same thing but for those 18 people randomized to drink water


### Example: Beer consumption and mosquito attraction to humans

Descriptives: Does there look to be a difference between the groups?

```{r, echo=F, fig.align='center', out.width="80%"}
library(ggplot2)

ggplot(mosq_data, aes(x = treatment, y = num_mosquitos)) +
  geom_boxplot(aes(fill = treatment)) + 
  theme_minimal(base_size = 15) +
  labs(y = "Number of mosquitos", x = "")
```



### Example: Beer consumption and mosquito attraction to humans
What type of variable is the exposure here?

What type of variable is the outcome?

How might we approach this with a parametric test?

### Example: Beer consumption and mosquito attraction to humans
```{r}
t.test(beer, water, alternative = "two.sided")
```
The average number of mosquitoes attracted to beer drinkers was 23.6 vs. 19.22 
attracted to water drinkers.  What is the p-value?

### Example: Beer consumption and mosquito attraction to humans

There is another way to perform this test. Consider the null hypothesis:

$$H_0: \mu_1 = \mu_2$$

If the two means are the same, then we would expect no difference between the 
number of mosquitoes attracted to beer drinkers vs. water drinkers. 



Watch this clip from 8:13-9:52: https://youtu.be/5Dnw46eC-0o?t=492.

- It shows the sampling distribution being built for this example under
the null hypothesis of no difference.
- It shows how the labels can be shuffled at random, and after each re-shuffling, 
the mean difference is computed and plotted on an evolving histogram.
- Then a vertical line is added at the **observed** value of the difference 
(based on the data from the sample).
- An observed value in the tails of the distribution implies that it is unlikely
to occur under the null hypothesis.

## The 'infer' package in R

### The `infer` package

The `infer` package is relatively new to the tidyverse (which includes `ggplot2`,
`readr`, `dplyr`, among others)

It is **awesome** because it interjects the steps of hypothesis testing directly
into the code. It also keeps things "tidy" meaning that the output is often
returned in a nice little data frame.

You can use `infer` to conduct permutation tests, but if you're interested you 
could also learn more [here](https://infer.netlify.com/) about doing all your
testing using this package.

Let's have a look!

### The `infer` package for permutation tests

First use the `infer` functions `specify()`, `hypothesize()`, `generate()`, 
and `calculate` to create the histogram of the sampling distribution for the 
mean difference:

```{r, warning=FALSE, message=FALSE}
null_distn <- mosq_data %>% 
  specify(response = num_mosquitos, explanatory = treatment) %>%
  hypothesize(null = "independence") %>%
  generate(reps = 1000, type = "permute") %>%
  calculate(stat = "diff in means", order = c("beer", "water"))
```


### The `infer` package for permutation tests
```
head(null_distn)
```


### The `infer` package for permutation tests

Then, use the `infer` function `visualize` to plot the sampling distribution, 
add a line at the observed mean difference, and shade the region corresponding
to the p-value:

```{r, fig.align='center', out.width="60%", warning=FALSE, message=FALSE}
null_distn %>% visualize(obs_stat = 23.6-19.22, direction = "two_sided")  
```

### The `infer` package for permutation tests

Finally, calculate the p-value by using the `get_pvalue()` function:

```{r}
null_distn %>% get_pvalue(obs_stat = 23.6-19.22, direction = "two_sided")
```
